Temporal Dynamics of Normalization Reweighting

### 2022-09-30 manuscript.tex
Minor changes to keep track of rejected trials.

\title{Temporal Dynamics of Normalization Reweighting}
\author{Daniel H. Baker, Daniela Marinova, Richard Aveyard, Lydia Hargreaves, Alice Renton,\\
Ruby Castellani, Phoebe Hall, Miriam Harmens, Georgia Holroyd, Beth Nicholson,\\
Emily Williams, Hannah M. Hobson \& Alex R. Wade}
\date{2022-09-11}

@@

Normalization itself is not a single process. Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation (Foley, 1994; Heeger, 1992), eye-of-origin (Legge, 1979; Sengpiel and Blakemore, 1994) and spatial position (Cannon and Fullenkamp, 1991; Petrov, 2005). At present there is evidence of normalization reweighting between stimuli with orthogonal orientations (Aschner et al., 2018), and adjacent spatial positions (Yiltiz et al., 2020). We anticipated that interocular suppression should also be subject to reweighting, but that there might be differences in the dynamics across suppressive pathways. This is plausible, given that suppression within and between the eyes has different spatiotemporal tuning (Meese and Baker, 2009), and dichoptic masking can be reduced by adapting to the mask (Baker et al., 2007; Li et al., 2005; Sengpiel and Vorobyov, 2005), unlike within-eye masking. Dynamic fluctuations in interocular suppression are a key feature of binocular rivalry (Alais et al., 2010; Wilson, 2003), for which autism-related differences have also been reported (Robertson et al., 2013).

@@

Our recent work has identified a potential autism-related difference in gain control using steady-state EEG methods. Vilidaite et al. (2018) measured contrast response functions for flickering stimuli in a large group of neurotypical adults, who were split by their autism quotient (AQ, see Baron-Cohen et al., 2001) score. The responses were comparable between high and low AQ groups at the flicker frequency of the stimulus. However at the second harmonic frequency (twice the flicker frequency), the high AQ group showed weaker responses than the low AQ group. This effect was replicated in an independent sample of 12 adults with an autism diagnosis (compared with neurotypical controls). Because second harmonic responses are caused by nonlinear interactions in the visual system (they are absent in a linear system), and divisive suppression is a nonlinearity, this potentially implicates differences in suppression in autism. Furthermore, there appears to be a developmental trajectory, as autistic children tested as part of the same study (Vilidaite et al., 2018) showed weaker responses at both the first and second harmonic frequencies.

@@

{\centering \includegraphics{temp/figures/Fig1}

@@

\caption{Summary of pilot analysis of data from Vilidiate et al. (2018). Panel (a) shows contrast response functions at the target frequency (7Hz, left) and the mask frequency (5Hz, right). Insets show the distribution of activity across the scalp, with points marking the electrodes over which signals were averaged (Oz, POz, O1 and O2). Panel (b) shows Fourier spectra for the single component stimuli and their combination (plaid). Note the strong second harmonic components at 14Hz and 10Hz. Panel (c) shows timecourses of frequency-locked responses to a single stimulus (black) and the plaid stimulus (blue), compared to baseline (grey). Panel (d) shows the timecourse of suppression at each frequency (7Hz, 5Hz, 14Hz, 10Hz) and their average (black curve). Points around y = 1.8 indicate a significantly increasing ratio. Error bars in panel (a) and shaded regions in panels (c,d) indicate ±1SE, and grey rectangles indicate the timing of stimulus presentation. The larger symbols in panel (a) indicate the conditions used for subsequent analyses.}\label{fig:Pilotdata}
\end{figure}

@@

Our initial reanalysis was promising, however the data were noisy despite the large sample size, because each participant contributed only 8 trials (88 seconds) to each condition. We therefore preregistered two new experiments (see \url{https://osf.io/4qudc}) to investigate these effects in greater detail. These had a similar overall design to the Vilidaite et al. (2018) study, with some small changes intended to optimise the study (see Methods). The key differences were that we used shorter trials (because there were few changes in the latter part of the trials shown in Figure \ref{fig:Pilotdata}d), and also focussed all trials into a smaller number of conditions, such that each participant contributed 48 repetitions (288 seconds of data) to each of 4 conditions.

@@

One important question is whether the dynamic increase in suppression can be explained by the stimulus onset transient. This is a possibility that cannot be ruled out for some of our data. For example, the steep increase in suppression in Figure \ref{fig:EEGdata}f has a similar timecourse to the onset transient in Figure \ref{fig:EEGdata}d.~However, there are also counterexamples where suppression continues to increase well beyond the first 1 second of stimulus presentation (e.g.~Figure \ref{fig:EEGdata}e). It is currently unclear why there appear to be such substantial differences between temporal frequency conditions, especially with such similar frequencies (5 and 7Hz). However the differences are relatively consistent across experiments. For example, 5Hz flicker produces a more gradual increase in suppression across all three data sets, compared with 7Hz flicker. These differences may be a consequence of visual channels with different temporal tuning interacting with the stimulation frequency, as well as any nonlinearities that govern suppression. Or there could be an asymmetry, whereby the relative temporal frequency between the two stimulus components affects the character of suppression (Liza and Ray, 2022). We hope that in future dynamic models of early vision (e.g. Groen et al., 2022) will be able to explain these differences.

@@

In Experiments 1 and 2, stimuli were presented using a ViewPixx LCD display device (VPixx Technologies, Canada) with a resolution of 1920 \(\times\) 1080 pixels and a refresh rate of 120Hz. The display was gamma corrected using a Minolta LS110 photometer. In Experiment 2, participants wore active stereo shutter glasses (NVidia 3D Vision 2) that were synchronised with the display using an infra-red signal. EEG data were collected using a 64-channel Waveguard cap, and were amplified and digitised at 1000Hz using an ANT Neuroscan system. Electrode impedance was maintained below 5k\(\Omega\), and referenced to a whole-head average.

@@

EEG data from Experiments 1 and 2 were first imported into Matlab using components of the EEGlab toolbox (Delorme and Makeig, 2004), and converted into a compressed ASCII format. Primary data analysis was then conducted using a bespoke \emph{R} script. In brief, we epoched each trial and extracted the average timecourse across four occipital electrodes (\emph{Oz}, \emph{POz}, \emph{O1} and \emph{O2}), and then calculated the Fourier transform of this average waveform. We excluded trials for which the Mahalanobis distance of the complex Fourier components exceeded 3 (for details see Baker, 2021). This resulted in \% of trials being excluded for Experiment 1, \% of trials for Experiment 2, and \% of trials for Experiment 3. Next we averaged the waveforms across all remaining trials, and calculated the Fourier transform in a 1-second sliding window to generate timecourses for each participant. We divided the timecourse for the target-only condition by the timecourse for the target + mask condition to produce a suppression ratio. These were then converted to logarithmic (dB) units for averaging, calculation of standard errors, and statistical comparisons. For display purposes we smoothed the timecourses using a cubic spline function, however all statistical comparisons used the unsmoothed data.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Groen2022}{}}%
Groen IIA, Piantoni G, Montenegro S, Flinker A, Devore S, Devinsky O, Doyle W, Dugan P, Friedman D, Ramsey N, Petridou N, Winawer J. 2022. Temporal dynamics of neural responses in human visual cortex. \emph{J Neurosci}. doi:\href{https://doi.org/10.1523/JNEUROSCI.1812-21.2022}{10.1523/JNEUROSCI.1812-21.2022}

@@

@@

>>>>

\title{Temporal Dynamics of Normalization Reweighting}
\author{Daniel H. Baker, Daniela Marinova, Richard Aveyard, Lydia J. Hargreaves, Alice Renton,\\
Ruby Castellani, Phoebe Hall, Miriam Harmens, Georgia Holroyd, Beth Nicholson,\\
Emily L. Williams, Hannah M. Hobson \& Alex R. Wade}
\date{2022-09-30}

@@

Normalization itself is not a single process. Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation (Foley, 1994; Heeger, 1992), eye-of-origin (Legge, 1979; Sengpiel and Blakemore, 1994) and spatial position (Cannon and Fullenkamp, 1991; Petrov, 2005). At present there is evidence of normalization reweighting between stimuli with orthogonal orientations (Aschner et al., 2018), and adjacent spatial positions (Yiltiz et al., 2020). We anticipated that interocular suppression should also be subject to reweighting, but that there might be differences in the dynamics across suppressive pathways. This is plausible, given that suppression within and between the eyes has different spatiotemporal tuning (Meese and Baker, 2009), and dichoptic masking can be reduced by adapting to the mask (Baker et al., 2007; Li et al., 2005; Sengpiel and Vorobyov, 2005), unlike within-eye masking. Dynamic fluctuations in interocular suppression are a key feature of binocular rivalry (Alais et al., 2010; Wilson, 2003), for which autism-related differences have also been reported (Robertson et al., 2013). By comparing monocular and dichoptic suppression, we are also able to isolate any contribution of early pre-cortical factors. This is because interocular suppression impacts in primary visual cortex, and therefore bypasses earlier retinal and subcortical stages of processing that may contribute to monocular suppression (Li et al., 2005).

@@

Our recent work has identified a potential autism-related difference in gain control using steady-state EEG methods. Vilidaite et al. (2018) measured contrast response functions for flickering stimuli in a large group of neurotypical adults, who were grouped according to their autism quotient (AQ, see Baron-Cohen et al., 2001) score. The responses were comparable between high and low AQ groups at the flicker frequency of the stimulus. However at the second harmonic frequency (twice the flicker frequency), the high AQ group showed weaker responses than the low AQ group. This effect was replicated in an independent sample of 12 adults with an autism diagnosis (compared with neurotypical controls). Because second harmonic responses are caused by nonlinear interactions in the visual system (they are absent in a linear system), and divisive suppression is a nonlinearity, this potentially implicates differences in suppression in autism. Furthermore, there appears to be a developmental trajectory, as autistic children tested as part of the same study (Vilidaite et al., 2018) showed weaker responses at both the first and second harmonic frequencies.

@@

{\centering \includegraphics{Figures/VilidaiteData}

@@

\caption{Summary of pilot analysis of data from Vilidiate et al. (2018). Panel (a) shows contrast response functions at the target frequency (7Hz, left) and the mask frequency (5Hz, right). Insets show the distribution of activity across the scalp, with points marking the electrodes over which signals were averaged (Oz, POz, O1 and O2). Panel (b) shows Fourier spectra for the single component stimuli and their combination (plaid). Note the strong second harmonic components at 14Hz and 10Hz. Panel (c) shows timecourses of frequency-locked responses to a single stimulus (black) and the plaid stimulus (blue), compared to baseline (grey). Panel (d) shows the timecourse of suppression at each frequency (7Hz, 5Hz, 14Hz, 10Hz) and their average (black curve). Points around y = 1.8 indicate a significantly increasing ratio. Error bars in panel (a) and shaded regions in panels (c,d) indicate ±1SE across N=100 participants, and grey rectangles indicate the timing of stimulus presentation. The larger symbols in panel (a) indicate the conditions used for subsequent analyses.}\label{fig:Pilotdata}
\end{figure}

@@

Our initial reanalysis was promising, however the data were noisy despite the large sample size (of N=100), because each participant contributed only 8 trials (88 seconds) to each condition. We therefore preregistered two new experiments (see \url{https://osf.io/4qudc}) to investigate these effects in greater detail. These had a similar overall design to the Vilidaite et al. (2018) study, with some small changes intended to optimise the study (see Methods). The key differences were that we used shorter trials (because there were few changes in the latter part of the trials shown in Figure \ref{fig:Pilotdata}d), and also focussed all trials into a smaller number of conditions, such that each participant contributed 48 repetitions (288 seconds of data) to each of 4 conditions.

@@

One important question is whether the dynamic increase in suppression can be explained by the stimulus onset transient. This is a possibility that cannot be ruled out for some of our data. For example, the steep increase in suppression in Figure \ref{fig:EEGdata}f has a similar timecourse to the onset transient in Figure \ref{fig:EEGdata}d.~However, there are also counterexamples where suppression continues to increase well beyond the first 1 second of stimulus presentation (e.g.~Figure \ref{fig:EEGdata}e). It is currently unclear why there appear to be such substantial differences between temporal frequency conditions, especially with such similar frequencies (5 and 7Hz). However the differences are relatively consistent across experiments. For example, 5Hz flicker produces a more gradual increase in suppression across all three data sets, compared with 7Hz flicker. These differences may be a consequence of visual channels with different temporal tuning interacting with the stimulation frequency, as well as any nonlinearities that govern suppression. Or there could be an asymmetry, whereby the relative temporal frequency between the two stimulus components affects the character of suppression (Liza and Ray, 2022). We hope to be able to model these effects in the future, for example by using dynamic models of early vision that incorporate time-lagged gain control (e.g. Zhou et al., 2019).

We did not observe clear differences in the timecourse between monocular and dichoptic suppression. This is important, because the dichoptic arrangement bypasses early stages of processing before the cortex (e.g.~the retina and lateral geniculate nucleus). It suggests that the dynamic increases in suppression occur in the cortex, consistent with our MEG data that find evidence of reweighting in V1 (see Fig \ref{fig:MEGdata}), and with previous neurophysiological work (Aschner et al., 2018). It is currently unclear whether these effects originate in V1, or might involve feedback from higher areas. The similarity between monocular and dichoptic effects also differs from work on adaptation to individual mask components. In both physiological (Li et al., 2005; \textbf{Sengpiel2006?}) and psychophysical (Baker et al., 2007) paradigms, adapting to a dichoptic mask reduces its potency, whereas adapting to a monocular mask has little or no effect. Normalization reweighting offers an explanation for why monocular masks presented in isolation do not adapt: if suppressive weights are determined by co-occurrence of stimuli, presentation of an isolated mask will have little effect. However this cannot explain the dichoptic adaptation effects without invoking additional binocular processes, such as competition between summing and differencing channels (e.g. May et al., 2012).

@@

In Experiments 1 and 2, stimuli were presented using a ViewPixx 3D LCD display device (VPixx Technologies, Canada) with a resolution of 1920 \(\times\) 1080 pixels and a refresh rate of 120Hz. The display was gamma corrected using a Minolta LS110 photometer. In Experiment 2, participants wore active stereo shutter glasses (NVidia 3D Vision 2) that were synchronised with the display using an infra-red signal. EEG data were collected using a 64-channel Waveguard cap, and were amplified and digitised at 1000Hz using an ANT Neuroscan system. Electrode impedance was maintained below 5k\(\Omega\), and referenced to a whole-head average.

@@

EEG data from Experiments 1 and 2 were first imported into Matlab using components of the EEGlab toolbox (Delorme and Makeig, 2004), and converted into a compressed ASCII format. Primary data analysis was then conducted using a bespoke \emph{R} script. In brief, we epoched each trial and extracted the average timecourse across four occipital electrodes (\emph{Oz}, \emph{POz}, \emph{O1} and \emph{O2}), and then calculated the Fourier transform of this average waveform. We excluded trials for which the Mahalanobis distance of the complex Fourier components exceeded 3 (for details see Baker, 2021). This resulted in 0.25\% of trials being excluded for Experiment 1, 4.51\% of trials for Experiment 2, and 4.51\% of trials for Experiment 3. Next we averaged the waveforms across all remaining trials, and calculated the Fourier transform in a 1-second sliding window to generate timecourses for each participant. We divided the timecourse for the target-only condition by the timecourse for the target + mask condition to produce a suppression ratio. These were then converted to logarithmic (dB) units for averaging, calculation of standard errors, and statistical comparisons. For display purposes we smoothed the timecourses using a cubic spline function, however all statistical comparisons used the unsmoothed data.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-May2012}{}}%
May KA, Zhaoping L, Hibbard PB. 2012. Perceived direction of motion determined by adaptation to static binocular images. \emph{Curr Biol} \textbf{22}:28--32. doi:\href{https://doi.org/10.1016/j.cub.2011.11.025}{10.1016/j.cub.2011.11.025}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Zhou2019}{}}%
Zhou J, Benson NC, Kay K, Winawer J. 2019. Predicting neuronal dynamics with a delayed gain control model. \emph{PLoS Comput Biol} \textbf{15}:e1007484. doi:\href{https://doi.org/10.1371/journal.pcbi.1007484}{10.1371/journal.pcbi.1007484}
